% vim: ft=tex ts=2 sw=2 foldlevel=2

\begin{thebibliography}{99}

	\bibitem{elfspec-sysv}
		\emph{Chapter 4: Object Files}, in
		\emph{System V Application Binary Interface Edition 4.1} (pages 44-72) \\
		The Santa Cruz Operation, AT\&T, The 88open Consortium \\
		\url{http://www.sco.com/developers/devspecs/gabi41.pdf} \\
		Accessed: May 3rd, 2015.

	% \bibitem{tis-elf}
	% 	\emph{Tool Interface Standard (TIS) Executable and Linking Format (ELF)
	% 	Specification, version 1.2} \\
	% 	Tool Interface Standard Committee \\
	% 	Edited May 1995. \\
	% 	\url{http://refspecs.linuxbase.org/elf/elf.pdf} \\
	% 	Accessed: May 4rd, 2015.

	\bibitem{dwarfspecv4}
		\emph{DWARF Debugging Information Format Version 4} \\
		DWARF Debugging Information Format Committee \\
		\url{http://dwarfstd.org/doc/DWARF4.pdf} \\
		Accessed: May 5th, 2015.

	\bibitem{debugdwarf}
		\emph{Introduction to the DWARF Debugging Format} \\
		Michael J. Eager \\
		\url{http://dwarfstd.org/doc/Debugging\%20using\%20DWARF-2012.pdf} \\
		Accessed: May 5th, 2015.

	\bibitem{howdebugworks}
		\emph{Part 3 -  Debugging Information}, in \emph{How Debuggers Work} \\
		Eli Bendersky \\
		\url{http://eli.thegreenplace.net/2011/02/07/how-debuggers-work-part-3-debugging-information/} \\
		Accessed: May 5th, 2015.

	\bibitem{tratt-dynamic-langs}
		\emph{Dynamically Typed Languages},
		in \emph{Advances in Computers} (volume 77, pages 149-184) \\
		Laurence Tratt \\
		Edited by Marvin V. Zelkowitz (July 2009) \\
		\url{http://tratt.net/laurie/research/pubs/html/tratt__dynamically_typed_languages/} \\
		Accessed: August 22nd, 2015.

	\bibitem{lua-pil}
		\emph{Programming in Lua} \\
		Roberto Ierusalimschy \\
		Lua.org, Third Edition (January 3rd, 2013), ISBN 859037985X.

	\bibitem{lua-pil-online}
		\emph{Programming in Lua} \\
		Roberto Ierusalimschy \\
		Lua.org, First Edition (December 2003), ISBN 8590379817. \\
		\url{http://www.lua.org/pil/contents.html}

	\bibitem{lua-about}
		\emph{About}, in \emph{Lua website} \\
		\url{http://www.lua.org/about.html} \\
		Checked April 20th, 2015.

	\bibitem{lua-manual}
		\emph{Lua 5.3 Reference Manual} \\
		Roberto Ierusalimschy, Luiz Henrique de Figueiredo,
		Waldemar Celes \\
		\url{http://www.lua.org/manual/5.3/manual.html} \\
		Accessed: April 29th, 2015.

	\bibitem{lua50-impl}
		\emph{The Implementatin of Lua 5.0} \\
		Roberto Ierusalimschy, Luiz Henrique de Figueiredo,
		Waldemar Celes \\
		Journal of Universal Computer Science 11 \#7 (2005) \\
		\url{http://www.jucs.org/jucs_11_7/the_implementation_of_lua}

	\bibitem{lj-ffi-api}
		\emph{ffi.* API Functions} \\
		Mike Pall \\
		\url{http://luajit.org/ext_ffi_api.html} \\
		Accessed: May 11th, 2015.

	\bibitem{luaffi}
		\emph{luaffi: Standalone FFI library for calling C functions from Lua}  \\
		James McKaskill \\
		\url{https://github.com/jmckaskill/luaffi} \\
		Accessed: May 11th, 2015.

	\bibitem{lj-ffi-semantic}
		\emph{FFI Semantics} \\
		Mike Pall \\
		\url{http://luajit.org/ext_ffi_semantics.html} \\
		Accessed: May 11th, 2015.

	\bibitem{opengroup-dlopen}
		\emph{dlopen - open a symbol table handle}, in \emph{The Open Group Base
		Specifications Issue 7, IEEE Standard 1003.1 2013 Edition} \\
		The Open Group, IEEE \\
		\url{http://pubs.opengroup.org/onlinepubs/9699919799/functions/dlopen.html}	\\
		Accessed: August 30th, 2015.

	\bibitem{scrumban-getting-started}
		\emph{Getting Started with Scrumban} \\
		\url{http://www.aboutscrumban.com/how-to-start-using-scrumban/} \\
		Accessed: May 5th, 2015.

	\bibitem{tap-spec}
		\emph{TAP Specification} \\
		Michael G. Schwern, Andy Lester \\
		\url{http://testanything.org/tap-specification.html} \\
		Accessed: May 4th, 2015.

	\bibitem{libdwarf-doc}
		\emph{A Consumer Library Interface to DWARF} \\
		David Anderson \\
		\url{https://github.com/Distrotech/libdwarf/blob/distrotech-libdwarf/libdwarf/libdwarf2.1.pdf} \\
		Accessed: May 11th, 2015.

	\bibitem{libdwarfp-doc}
		\emph{A Producer Library Interface to DWARF} \\
		David Anderson \\
		\url{https://github.com/Distrotech/libdwarf/blob/distrotech-libdwarf/libdwarf/libdwarf2p.1.pdf} \\
		Accessed: May 11th, 2015.

	\bibitem{ninja-manual}
		\emph{Ninja documentation} \\
		Evan Martin \\
		\url{http://martine.github.io/ninja/manual.html} \\
		Accessed: May 11th, 2015.

	\bibitem{gnumake-manual}
		\emph{GNU Make: A Program for Directing Recompilation} \\
		Free Software Foundation, ISBN 1-882114-83-3. \\
		\url{https://www.gnu.org/software/make/manual/make.pdf} \\
		Accessed: May 13th, 2015.

	\bibitem{uthash-guide}
		\emph{uthash User Guide} \\
		Troy D. Hanson \\
		\url{https://troydhanson.github.io/uthash/userguide.html} \\
		Accessed: May 15th, 2015.

	\bibitem{swig3doc}
		\emph{SWIG 3.0 Documentation} \\
		\url{http://swig.org/Doc3.0/SWIGDocumentation.html} \\
		Accessed: June 25th, 2015.

	\bibitem{lusers-BindingCodeToLua}
		\emph{Binding Code To Lua}, in \emph{Lua-Users Wiki} \\
		\url{http://lua-users.org/wiki/BindingCodeToLua} \\
		Accessed: June 25th, 2015.

	\bibitem{js-raceforspeed}
		\emph{The JavaScript engine family tree},
		in \emph{The race for speed, part 1}  \\
		John Dalziel, CreativeJS \\
		\url{http://creativejs.com/2013/06/the-race-for-speed-part-1-the-javascript-engine-family-tree/} \\
		Accessed: August 16th, 2015.

	\bibitem{gobject-introspection}
		\emph{GObject Introspection}, in \emph{GNOME Wiki} \\
		\url{https://wiki.gnome.org/Projects/GObjectIntrospection} \\
		Accessed: August 15th, 2015.

	\bibitem{lj-perf1}
		\emph{LuaJIT performance}, in \emph{lua-l mailing list, August 2009} \\
		Mike Pall \\
		\url{http://lua-users.org/lists/lua-l/2009-08/msg00151.html} \\
		Accessed: August 22nd, 2015

	\bibitem{autotools-history}
		\emph{The First Configure Programs},
		in \emph{Autoconf, Automake, and Libtool} \\
		Gary V. Vaughan, Ben Elliston, Tom Tromey and Ian Lance Taylor \\
		New Riders Publishing (October 2000; updated February 2006). \\
		\url{https://www.sourceware.org/autobook/autobook/autobook_8.html} \\
		Accessed: August 27th, 2015.

\end{thebibliography}

